Comparison of dimensionality reduction schemes for dervative-free global optimization algorithms

### 2018-05-13 2018/YSC/ysc_paper.tex
YSC: update abstract and introduction

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Comparison of dimensionality reduction schemes for direct global optimization
algorithms}

@@

\keywords{Global optimization, Dimension reduction,
Multidimensional multiextremal optimization, Global search algorithms}

\begin{abstract}
This work considers a parallel algorithms for solving multi-extremal optimization problems.
Algorithms are developed within the framework of the information-statistical approach and
implemented in a parallel solver Globalizer. The optimization problem is solved by reducing
the multidimensional problem to a set of joint one-dimensional problems that are solved in
parallel. Five types of Peano-type space-filling curves are employed to reduce dimension. The
results of computational experiments carried out on several hundred test problems are discussed.
\end{abstract}

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

In the present paper, the parallel algorithms for solving the multiextremal optimization problems
are considered. In the multiextremal problems, the opportunity of reliable estimate of the global
optimum is based principally on the availability of some information on the function known
{\textit a priori} allowing relating the probable values of the optimized function to the known
values at the points of performed trials. Very often, such an information on the problem being
solved is represented in the form of suggestion that the objective function $\varphi(y)$ satisfies
Lipschitz condition with the constant $L$ not known a priori (see, for example,
\cite{Jones2009,Gablonsky2001,Evtushenko2013}). At that, the objective function could be defined by a
program code i. e. could represent a ``black-box''-function. Such problems are presented in the
applications widely (problems of optimal design of objects and technological processes in
various fields of technology, problems of model fitting according to observed data in scientific
research, etc.).

Many methods destined to solving the problems of the class specified above reduce the solving
of a multidimensional problem to solving the one-dimensional subproblems implicitly (see, for
example, the methods of diagonal partitions \cite{Sergeyev2006,SergeyevKvasov2015} or
simplicial partitions \cite{Zilinskas2008,Zilinskas2014}). In the present work, we will use the
approach developed in Lobachevsky State University of Nizhni Novgorod based on the idea of
the dimensionality reduction with the use of Peano space-filling curves $y(x)$ mapping the
interval $[0,1]$ of the real axis onto an $n$-dimensional cube continuously and unambiguously.

Several methods of constructing the evolvents approximating the theoretical Peano curve have
been proposed in \cite{strongin1978,Strongin1992,Goryachih2017,Gergel2009}. These
methods were implemented in the Globalizer software system \cite{globalizerSystem}. The goal
of the present study was comparing the properties of the evolvents and the selecting the most
suitable ones for the use in the parallel global optimization algorithms.

%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbb{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function
\(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbb{R}^N\)
space.
\par
Usually, the objective function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is time-consuming.
%As a result, the overall time of solving the optimization
%problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
%It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
%is highly important, since an estimate of the global minimum can be constructed on the
%basis of a finite number of computed values of the optimized function only in this case .

>>>>

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Comparison of dimensionality reduction schemes for dervative-free global optimization
algorithms}

@@

\keywords{Global optimization, Dimension reduction, Derivative-free algorithms, Global search algorithms}

\begin{abstract}
  A common approach to solving global optimization problems is to use
  univariate optimization algorithms in combination with dimensional reduction schemes.
  The paper considers five types of Peano-like space-filling curves (evolvents), which are used
  to reduce the dimension in the derivative-free algorithm of global optimization. The algorithm
  is univariate and developed within the framework of the information-statistical approach.
  This work is the first one, where convergence rates and implementations details
  of these five evolvents are considered together and directly compared.
\end{abstract}

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
In the present paper, the algorithms for solving the multiextremal optimization problems
are considered. In the multiextremal problems, the opportunity of a reliable estimate of the global
optimum is based principally on the availability of some information on the function known
{\textit a priori} allowing relating the probable values of the optimized function to the known
values at the points of performed trials. Very often, such an information on the problem being
solved is represented in the form of suggestion that the objective function $\varphi(y)$ satisfies
Lipschitz condition with the constant $L$ not known a priori (see, for example,
\cite{Jones2009,Gablonsky2001,Evtushenko2013}). At that, the objective function could be represented by a ``black-box''-function.
Many methods destined to solving the problems of the class specified above reduce the solving
of a multidimensional problem to solving the one-dimensional subproblems implicitly (see, for
example, the methods of diagonal partitions \cite{Sergeyev2006,SergeyevKvasov2015} or
simplicial partitions \cite{Zilinskas2008,Zilinskas2014}). In the present work, we will use the
approach developed in Lobachevsky State University of Nizhni Novgorod based on the idea of
the dimensionality reduction with the use of Peano space-filling curves $y(x)$ mapping the
interval $[0,1]$ of the real axis onto an $n$-dimensional cube continuously and unambiguously.

In recent years several methods of constructing Peano-type space-filling curves (\textit{evolvents})
have been proposed \cite{strongin1978,Strongin1992,Goryachih2017,Gergel2009}. Some of them were
succesful used to build large-scale parallel algorithms \cite{stronginGergelBarkalovParGO}.
Presented paper numerically compares different evolvents in terms of convergence rates and
computational efficiency. The comparison is done by solving sets consisting of hundreds of test problems and collecting
different characteristics of convergence. As a result, practical recommendations about
the usage of each evolvent type are given. For example, some evolvents are suitable for parallel computations,
while others are quite useful in case of sequential solving of problems with time-consuming objective functions.

%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
univariate algorithm of global search \cite{strongin1978}, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbb{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function
\(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbb{R}^N\)
space.
\par
Usually, the objective function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is time-consuming.
%As a result, the overall time of solving the optimization
%problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
%It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
%is highly important, since an estimate of the global minimum can be constructed on the
%basis of a finite number of computed values of the optimized function only in this case .
